Basics of Hardware Cache Optimization

### 2017-06-25 tex/com.tex
Expand, tweak, add some code but don't use it
I expected this code to perform better...

The algorithm is \alts{by definition, technically} already cache-oblivious since it
doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
with a divide-and-conquer approach.  The idea is to recursively divide the input matrix
\(D\) into two equal-sized submatrices along the greater dimension.  If \(m \geq n\) (more
rows than columns), let
\begin{equation*}
   D = \begin{bmatrix}
      D_1\\
      D_2
   \end{bmatrix}
\end{equation*}
and use \(D^\mathsf{T} = \begin{bmatrix} D_1^\mathsf{T} & D_2^\mathsf{T} \end{bmatrix}\)
to compute the transpose.  If \(m < n\), analogously split \(D\) vertically.  Eventually,
the submatrices will become small enough so that pairs of input and output submatrices fit
into cache at the same time, at which point it doesn't matter in what order we access the
elements.\footnote{%
   No cache line will have to be evicted once loaded.
}
The recursion continues all the way down to \(1 \times 1\) submatrices, but this doesn't
change the theoretical analysis.  According to
\textcite[theorem 2 and 3, \ppno~19--21]{coa-thesis}
% \textcite[theorem 2 and 3, \ppno~19\psqq]{coa-thesis}
this algorithm incurs an optimal amount of \(\Theta(1 + mn / B)\) cache misses.
% This reduces the amount of cache misses to \(\Theta(1 + \nicefrac{mn}{B})\) according to
% \textcite[theorem 2, \ppno~19\psq]{coa-thesis}.

In practice, the new algorithm performs as badly as one might expect from this excessive
use of recursion.  \Cref{fig:xpose-speedup} gives the speedup compared to the
straightforward implementation, which is usually faster.

% The code may only be 72 characters wide (75 when counting the indentation that will be
% stripped).  Note that the array arguments will decay to pointers.
\begin{minted}[autogobble, mathescape]{c}
   // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
   void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
      int num_rows = 1 + I[1] - I[0];
      int num_cols = 1 + J[1] - J[0];
      if (num_cols == 1 && num_rows == 1) {
         E[J[0]][I[0]] = D[I[0]][J[0]];
      } else if (num_cols <= num_rows) {
         // Horizontally slice D into two submatrices and recurse.
         transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
         transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
      } else { /* Vertically slice D analogously... */ }
   }
\end{minted}
% // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
% // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
% // Transpose the submatrix $D\left[...\right]$

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \caption{TODO}
   \label{fig:xpose-speedup}
\end{figure}

>>>>

The algorithm is \alts{by definition, technically} already cache-oblivious since it
doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
with a divide-and-conquer approach.  The idea is to recursively divide the input matrix
\(D\) into two equal-sized submatrices along the greater dimension.  If \(m \geq n\) (more
rows than columns), let
\begin{equation*}
   D = \begin{bmatrix}
      D_1\\
      D_2
   \end{bmatrix}
\end{equation*}
and use \(D^\mathsf{T} = \begin{bmatrix} D_1^\mathsf{T} & D_2^\mathsf{T} \end{bmatrix}\)
to compute the transpose.  If \(m < n\), analogously \alts{slice, split} \(D\) vertically.
Eventually, the submatrices will become small enough so that pairs of input and output
submatrices fit into cache at the same time, at which point it doesn't matter in what
order we access the elements.\footnote{%
   No cache line will have to be evicted once loaded.
}
The recursion continues all the way down to \(1 \times 1\) submatrices, but this doesn't
change the theoretical analysis.  According to
\textcite[theorem 2 and 3, \ppno~19--21]{coa-thesis}
% \textcite[theorem 2 and 3, \ppno~19\psqq]{coa-thesis}
this algorithm incurs an optimal amount of \(\Theta(1 + mn / B)\) cache misses.
% This reduces the amount of cache misses to \(\Theta(1 + \nicefrac{mn}{B})\) according to
% \textcite[theorem 2, \ppno~19\psq]{coa-thesis}.
\Cref{lst:xpose} shows my implementation in C.

% FIXME: get rid of this `center` hack and get the vertical spacing around listings with
% captions to work in a better way.
\begin{center}
   % The code may only be 72 characters wide (78 when counting the indentation that will
   % be stripped).  Note that the array arguments will decay to pointers.
   \begin{minted}[autogobble, mathescape]{c}
      // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
      void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
         int num_rows = 1 + I[1] - I[0];
         int num_cols = 1 + J[1] - J[0];
         if (num_cols == 1 && num_rows == 1) {
            E[J[0]][I[0]] = D[I[0]][J[0]];
         } else if (num_cols <= num_rows) {
            // Horizontally slice D into two submatrices and recurse.
            transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
            transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
         } else { /* Vertically slice D analogously... */ }
      }
   \end{minted}
   % // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
   % // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
   % // Transpose the submatrix $D\left[...\right]$
   \captionof{listing}{\acrshort{com}-Optimal Matrix Transposition}
   \label{lst:xpose}
\end{center}

% In practice, the new algorithm performs as badly as one might expect from this excessive
% use of recursion.
In practice, the new algorithm performs worse than the straightforward one for a lot of
matrix sizes before it eventually pulls ahead.
\alts{
   {\Cref{fig:xpose-speedup} gives the speedups for square matrices of various sizes},
   {\Cref{fig:xpose-speedup} gives the speedup compared to the straightforward
   implementation\x{, which is usually faster}},
}.
The problem may be the excessive use of recursion.

Kumar works around this by \say{[stopping] the recursion when the problem size becomes
less than a certain block size and then [using] the simple for loop implementation inside
the block}~\cite[199--201]{afmh}.  This seems to \alts{compromise, jeopardize} the
\alts{cache-obliviousness, cache-oblivious nature} of the algorithm.

Perhaps cache-oblivious algorithms should be seen as a starting point for further
optimizations that \alts{explicitly use the available information about the memory
hierarchy of the machine a program runs on, take the properties of the target machine into
account}.

\x{
Another approach may be to
\alts{
   {slice the matrix so that the split dimension becomes an even power of \(2\) for one of
   the submatrices},
   {\alts{split, divide} the matrix so that one of the resulting submatrices has a size
   that is an even power of \(2\) along the dimension we split}
}.
\alts{
   {Once a square submatrix whose dimensions are an even power of \(2\) is obtained, the
   identical order of traversal that would result from the recursive algorithm can be
   reproduced with a single loop.},
   {Once a submatrix whose dimensions both are an even power of \(2\) is obtained, the
   identical order of traversal that would result from the recursive algorithm can be
   reproduced with loops relatively easily.},
}
}

\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Speedup Achieved by \acrshort{com}-Optimal Matrix Transposition]{Speedup
   Achieved by \acrshort{com}-Optimal Matrix Transposition; the CPU time used by the
   straightforward matrix transposition algorithm divided by that used by the optimal one}
   \label{fig:xpose-speedup}
\end{figure}

### 2017-06-25 tex/list-vs-vector.tex
Create more slides
Minor tweaks and additions
Improve spacing between listings and captions
Expand, tweak, add some code but don't use it
I expected this code to perform better...

\begin{center}
   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}
   \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
   \label{lst:ithare}
\end{center}

@@

\begin{figure}
   \centering
   \input{tex/oo-picture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Array of Pointers]{Array of Pointers; the array is compact but the actual
   objects may \alts{be scattered across memory pretty randomly, not be laid out
   sequentially in memory}.}
   \label{fig:true-oo}
\end{figure}

\subsubsection{Unrolled List}

>>>>

\begin{center}
   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}%
   \captionof{listing}[Compute the Sum of a Container]{Compute the Sum of a Container;
   adapted from \textcite{bigos}}
   \label{lst:ithare}
\end{center}

@@

\begin{figure}
   \centering
   \input{tex/graphics/oo-picture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Array of Pointers]{Array of Pointers; the array is compact but the actual
   objects may \alts{be scattered across memory pretty randomly, not be laid out
   sequentially in memory}.}
   \label{fig:true-oo}
\end{figure}

% \subsubsection{Unrolled List}
